Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-06 analysis/paper/paper.Rmd
minor edits

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic transition. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic transition? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.
